What is happening on Twitter? Tweet analysis and student research projects

### 2020-05-29 Rmd/abstract.Rmd
updated ms
revised abstract. added section numbers

We describe methods for collecting tweets from Twitter. With these methods, we 
curate a large collection of tweets.
We present methods for social media event detection through text analysis of tweets. 
We describe our use of tweets in mentored undergraduate research. 
We also offer other directions for other text analysis of tweets.

>>>>

We present a freely available, large, rich data set from Twitter and describe how we collected 
it. We then describe mentored student research that uses tweets to address novel research
questions. We use backward design principles to develop learning objectives
for student researchers. We conduct formative and summative assessments of student learning.
To illustrate the value of our data set, we 
present details of one student project. We conclude by discussing future research 
directions.

### 2020-05-29 Rmd/jse-special-issue-2020.Rmd
revised abstract. added section numbers

---
title: "What is happening on Twitter? Tweet analysis and undergraduate research projects"
author: "Frederick J. Boehm and Bret M. Hanlon"
date: "June 15, 2020"
params:
  lastmod: !r lubridate::now()
bibliography: jse.bib
csl: chicago-annotated-bibliography.csl
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
output: 
  pdf_document:
    latex_engine: xelatex
    toc: FALSE
    number_sections: FALSE
    includes:
      in_header: head.tex
  bookdown::word_document2:
    toc: FALSE
---

@@

```{r child = "abstract.Rmd", eval = FALSE}
```

## Introduction

@@

## Preliminary research mentoring considerations

@@

### Student experience with data analysis software

@@

### Student research interests and goals

@@

### Time period

@@

## Methods

### Collecting tweets over time

@@

### Querying Twitter API to get complete tweets

@@

From our collection of tweets, we chose to analyze those sent on five consecutive 
days from May 10, 2020 to May 14, 2020. To minimize the computing requirements, we 
limited our study to tweets sent during a five-minute period (12:00pm to 12:05pm Eastern time) every day.
We then submitted API queries to Twitter
to get the full content of tweets, including the tweet text.
In supplementary files, we provide the R code that we used to 
query the Twitter API to obtain full tweet content.

### Tweet structure

```{r, echo = FALSE, message = FALSE}
data_dictionary <- readr::read_csv("../data/tweets-data-dictionary.csv")
```

@@

### Parsing tweet text

@@

### Latent Dirichlet allocation

Latent Dirichlet allocation is a statistical method for inferring latent (unobservable) topics (or themes)
from a corpus (or collection) of documents. We pretend that
there's an imaginary process for creating documents in the corpus. For each document,
we choose a discrete distribution over topics. For example, some tweets from Memorial Day may refer to the holiday. This may constitute one topic in the corpus. 
Having chosen a distribution over topics, we then select document words by first 
drawing a topic from the distribution over topics, then drawing a word from the
chosen topic. 
Thus, a topic is technically defined as a distribution over words in a fixed vocabulary (or collection of words) [@blei2003latent].

### Study design

We sought to validate our hypothesis that we could detect a major social 
media event by examining tweet topic content at distinct time periods. 
As a proof of principle of our event detection strategy, we chose to 
analyze tweets before, during, and after Memorial Day 2020 (May 25, 2020). We fitted latent Dirichlet allocation models for each of three
distinct five-minute periods. The first period began at noon Eastern time on May 24, 2020. Subsequent time periods started 
24 and 48 hours later. 
We defined each time period to be a single collection, or corpus, of tweets. We then fitted latent 
Dirichlet allocation models to each corpus.

One challenge in latent Dirichlet allocation modeling is the choice of the 
number of topics per corpus. To address this, we fitted each corpus to models

We used several criteria to evaluate latent Dirichlet allocation model fits, with 
emphasis on choosing a small number of topics per corpus.
Our strategy involved both visualization and more quantitative approaches to model 
evaluation. For every model, we created one word cloud per topic.

We then inspected topic contents at each of the four time points.

### Results

```{r, fig.cap = "Top terms for LDA model from May 10, 2020 (Mother's Day)"}
knitr::include_graphics("../results/beta-2020-05-10.png")
```

```{r, fig.cap = "Top terms for LDA model from May 11, 2020"}
knitr::include_graphics("../results/beta-2020-05-11.png")
```

## Acknowledgements

@@

## Supplementary Material

## References

@@

## Appendix 1: Tweet data dictionary

@@

## Appendix 2: R analysis code

>>>>

---
title: "What is happening on Twitter? Tweet analysis and student research projects"
author: "Frederick J. Boehm and Bret M. Hanlon"
date: "June 15, 2020"
params:
  lastmod: !r lubridate::now()
bibliography: jse.bib
csl: chicago-annotated-bibliography.csl
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
output: 
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: FALSE
    number_sections: TRUE
    includes:
      in_header: head.tex
  bookdown::word_document2:
    toc: FALSE
---

@@

```{r child = "abstract.Rmd", eval = TRUE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introduction

@@

# Preliminary research mentoring considerations

@@

## Student experience with data analysis software

@@

## Student research interests and goals

@@

## Time period

@@

# Methods

## Collecting tweets over time

@@

## Querying Twitter API to get complete tweets

@@

From our collection of tweets, we chose to analyze those sent on three consecutive 
days from May 24, 2020 to May 26, 2020. We wanted to see if we could use text analysis 
tools to detect a transient change in topics for Memorial Day (May 25, 2020).

To minimize the computing requirements, we 
limited our analysis to tweets sent during a five-minute period (12:00pm to 12:05pm Eastern time) every day. However, our methods are appropriate for much larger data sets.
We then submitted API queries to Twitter
to get the full content of tweets, including the tweet text.
In supplementary files, we provide the R code that we used to 
query the Twitter API to obtain full tweet content.

## Tweet structure

```{r, echo = TRUE, message = FALSE}
data_dictionary <- readr::read_csv("../data/tweets-data-dictionary.csv")
```

@@

## Parsing tweet text

@@

## Latent Dirichlet allocation

Latent Dirichlet allocation is a statistical method for inferring latent (unobservable) topics (or themes)
from a large corpus (or collection) of documents [@blei2003latent].
We pretend that there's an imaginary process for creating documents in the corpus. For each document,
we choose a discrete distribution over topics. For example, some tweets from Memorial Day may 
refer to the holiday. This may constitute one topic in the corpus. 
Having chosen a distribution over topics, we then select document words by first 
drawing a topic from the distribution over topics, then drawing a word from the
chosen topic. 
Thus, a topic is technically defined as a distribution over words in a fixed vocabulary (or collection of words) [@blei2003latent].

## Study design

We sought to validate our hypothesis that we could detect a social 
media event by examining tweet topic content at distinct time periods. 
As a proof of principle of our event detection strategy, we chose to 
analyze tweets before, during, and after Memorial Day (May 25, 2020). We fitted latent Dirichlet allocation models for each of three
distinct five-minute periods. The first period began at noon Eastern time on May 24, 2020. Subsequent time periods started 
24 and 48 hours later. 
We defined each time period to be a single collection, or corpus, of tweets. We then fitted latent 
Dirichlet allocation models to each corpus.

# Results

We identified the top ten terms for each of ten topics in our models. We plotted 
the within-topic word probabilities as bar graphs.

```{r, fig.cap = "Top terms for LDA model from May 24, 2020"}
knitr::include_graphics("../results/beta-2020-05-24.png")
```

```{r, fig.cap = "Top terms for LDA model from May 25, 2020 (Memorial Day)"}
knitr::include_graphics("../results/beta-2020-05-25.png")
```

```{r, fig.cap = "Top terms for LDA model from May 26, 2020"}
knitr::include_graphics("../results/beta-2020-05-26.png")
```

Assigning meaning to topics is an active research area [@chang2009reading]. Since
our interest is in the transient appearance of a new topic, we don't attempt to
assign meaning to topics in our models.

We see that topic 7 from May 25 has several words that suggest Memorial Day: 
memorial, remember, honor, country. A similar topic is not seen on May 24 or May 26.

Some topics persist, possibly with distinct word probabilities, across the three days. For example, we see that President Trump features prominently in all three models.

We also note, on May 26, topic 10, which reflects discussion of the Amy Cooper Central Park incident.

We found evidence that latent Dirichlet allocation summarizes elements of 
large collections of Twitter conversations.

# Assessment of learning, exploring more advanced topics, and concluding remarks

## Assessment of learning

We examined student learning through both formative and summative assessments.
We conducted formative assessments through weekly discussions with students.
In these discussions, we developed action items to advance research progress and overcome challenges.

We summatively assessed student achievement at the end of the academic year. Both
students wrote a thesis and presented a poster to our statistics department. 
We asked questions at the poster session to probe student understanding and critically
evaluated the theses.

In future iterations, we will use a written rubric to grade student theses. We'll
share the rubric with our students at the start of the academic year.

## Exploring more advanced topics

Twitter data over time offers a wealth of potential research projects. Supplementing 
tweets with public data from other sources multiplies the possibilities. For example, 
one of our two students supplemented tweets with daily stock market index prices. 
She studied sentiment of finance-related tweets and daily stock market index
closing prices.

Latent Dirichlet allocation modeling and related methods are a major research area in the 
quantitative social sciences. Advanced students with interest in statistical computing might
compare inferential methods for topic models. Those with interests in event detection
and time series analysis could build on the findings of our student by explicitly accounting
for topic evolution with dynamic topic models [@blei2006dynamic].

## Concluding remarks

Tweet collections over time are a rich, large, authentic data set that offer 
many opportunities for student research projects. We provided 
instructions to enable readers to 
establish their own tweet collections. We also presented details for one 
mentored research project that made use of our stored tweets.

Tweet analysis gives students practical experience in the data science process of 
formulating a research question, gathering data to address it, summarizing the 
data, visualizing results, and communicating findings.

# Acknowledgements

@@

# References

@@

# Appendix 1: Tweet data dictionary

@@

# Appendix 2: R analysis code
